---
title: "lab2_homework"
author: "Scott Markley"
date: "February 16, 2015"
output: html_document
---

```{r load libraries, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(magrittr)
```

# Part 1. Analysis of El Malpais data

The tree ring data suggest that precipitation in certain months has a stronger relationship with annual tree ring width than precipitation in other months.

  The summer months experienced the greatest amount of rainfall during the study period, ranging from slightly under 2 inches in September to over 2.5 inches in both July and August, as seen in the "Precipitation by Month" bar chart pictured below. According to the same chart December and October were the only other two months with precipitation at or above one inch, while the rest of the months hovered between about 0.5 inches (May) to about 0.8 inches (January, February, and March). Tree ring width is most positively correlated with January precipitation, yielding a correlation coefficient of slightly over 0.3, as seen in red in the "correlation between RWI & Precipitation by Month" plot below. This plot also shows that following January, February precipitation also has a relatively high correlation with tree ring width. Rounding out the top six behind January and February in order are May, April, July, and then March, though each have fairly weak correlation coefficients of about 0.24 to 0.28. Additionally, from the same chart (shown in blue), the correlation is even stronger for January, February, and March when calculating r values with lagged precipitation data. 
  Overall, linear correlation is an appropriate way to measure the relationship between monthly precipitation and tree ring width. As the "Annual RWI vs. January Precipitation" and the "Annual RWI vs. August Precipitation" scatterplots show, this method does yield some useful results (i.e. an approximate positive association exists between January precipitation and RWI and no discernable relationship exists between August precipitation and RWI). However, for January, which has the the highest correlation coefficient of all the months, the relationship is only slightly positive, and as evidenced from the scatterplot, a good bit of variation exists.
  Finally, with regards to the last question (which I am not sure if I am interpreting correctly), according to the "Correlation between RWI & Precipitation by Month" plot, I would choose the 12 month period preceding the current year. In this plot, the lagged year r values appear to be, on average, stronger than the current year r values.

```{r}
load("C:/Users/Owner/Desktop/UT_S15/GEOG415/snmarkley/homework2/tidy.precip.R")
#Bar chart
tidy.precip %>% group_by(month) %>%
    summarize(precip=mean(precip, na.rm=TRUE)) %>%
    ggplot(data=., aes(month, precip)) + geom_bar(stat='identity') + 
    labs(x = 'Month', y = 'Precipitation (in)') +
    ggtitle('Precipitation by Month: 1895 to 2014')

```

```{r}
load("C:/Users/Owner/Desktop/UT_S15/GEOG415/snmarkley/homework2/rwi.precip.R")
##Scatterplots
#January
rwi.precip %>% filter(month=='Jan') %>%
     ggplot(data=., aes(precip,rwi)) + geom_point(stat='identity') +
     labs(x='January Precipitation (in)', y='Annual RWI') +
     geom_smooth(method=lm, se=FALSE) +
     ggtitle('Annual RWI vs. January Precipitation')
     
  #August
rwi.precip %>% filter(month=='Aug') %>%
     ggplot(data=., aes(precip,rwi)) + geom_point(stat='identity') +
     labs(x='August Precipitation (in)', y='Annual RWI') +
     geom_smooth(method=lm, se=FALSE) +
     ggtitle('Annual RWI vs. August Precipitation')

```
 
```{r}
##Plots
# I collaborated with Kyle in order to plot both current and lagged year on the same plot.
cor <- rwi.precip %>% group_by(month) %>% summarize(r=cor(rwi, precip, use='complete.obs'))
cor.lag <- rwi.precip %>% group_by(month) %>% summarize(r = cor(rwi, lag(precip),use='complete.obs'))
cor.both <- left_join(cor, cor.lag, by='month')
colnames(cor.both) <- c("month", "cor", "cor.lag")
ggplot(data=cor.both) + geom_point(aes(x=month, y=cor, color='Current year')) + 
                        geom_point(aes(x=month, y=cor.lag, color='Lagged year')) + 
                        labs(color='Correlation of:') + 
                        labs(x='Precipitation by Month', y= 'Correlation with RWI (r-value)') +
                        geom_abline(intercept=0, slope=0) + 
                        ggtitle('Correlation between RWI & Precipitation by Month: \n Current year & lagged year')

```
 

# Part 2: Getting the interpretation of correlation correct.

1. The second statement is more correct. Since a positive relationship between annual tree ring width and January precipitation would suggest wetter Januarys correlating with wider tree ring widths, one would conversely expect dryer Januarys to correlate with narrower tree ring widths. With regards to the first statement, Januarys are not usually wetter than the average month, and thus January precipitation relative to other months is not a sufficient explanation for why January precipitation levels positively correlate with annual tree ring width.

2.  The correlation would be negative because as the age of the car goes up, the price would (in most cases) go in the opposite direction. Taking antique cars into account would change that relationship though, since they typically go for higher prices. Then, as the age of cars goes up, their prices would go down until a certain point (50 years?) and then start going up again if the owner takes proper care of the vehicle.

3. False, the correlation coefficient is a standardized measure of the linear relationship between two variables that ranges from -1 to +1. It does not say what percentage of points are highly correlated.


4. No, this study does not provide a fair estimate for this association because the researcher uses state-level aggregates, making her analysis susceptible to the modifiable areal unit problem. Because the correlation between household income and rates of teenage pregnancy is 0.4 at the state level does not mean that relationship would hold at the individual level or even at different spatial scales.

5. GPA in freshman year and sophomore year in college would likely have a correlation coefficient of 0.6. This is because the grades of individual college students are likely to be very similar between years (i.e. positive correlation), although grades may not be exactly the same because of changes in course difficulty, students' adaptability to the college environment, changes in student attitudes toward education, etc. Therefore, I would expect a fairly high correlation, but not a perfect or near perfect correlation.
  GPA in freshman year and senior year in college would likely have a correlation coefficient of 0.3. There is probably some positive association between student performances at the beginning and at the end of college (e.g. students who earn high grades their freshman years are usually more likely to earn high grades their senior years). However, since more time has based between freshman and senior years than between freshman and sophomore years, I would expect that association to not be as strong.
  Length and weight of two-by-four boards would have a correlation coefficient close to 0.95. This is because as length increases, weight would almost definitely also increase since the dimensions are proportional. 

6. Probably not. Spending more hours watching television suggests that one will be foregoing opportunities to read, adversely affecting his/her reading ability. In other words, since someone is watching more television, he/she has less time to dedicate to reading and will therefore likely not score as well on reading test as his/her peers that watch less television.

